AppPAL for Android

### 2016-01-08 paper.tex
final version

There has been a great amount of work on developing app analysis tools for Android.
Tools such as Stowaway~\cite{Felt:2011kj} detect over-privileged apps.
TaintDroid~\cite{Enck:2010uw} and FlowDroid~\cite{Arzt:2014kf,Li:2015wo} can do taint and control flow analysis; sometimes even between app components.
Other tools like QUIRE~\cite{Bugiel:2012ui} can find privilege escalation attacks between apps.
ScanDAL~\cite{Kim:2012vt} and SCanDroid~\cite{Fuchs:2009vi} help detect privacy leaks.
Appscopy~\cite{Feng:kPGZr_ja} searches for specific kinds of malware.
Tools like DroidRanger~\cite{Zhou:2012tb} scan app markets for malicious apps.
Various tools such as AppGuard~\cite{Backes:2012vm}, Dr.~Android~\&~Mr.~Hide~\cite{Jeon:2012ki} or AppFence~\cite{Hornyack:2011wq} can control the permissions or data an app can get.
MalloDroid~\citep{Fahl:2012dj} looks for apps configured to use SSL incorrectly (for instance by not verifying hostnames or certificates).

@@

Other access control languages, such as XACML~\cite{Oasis:1M71_oFS}, could also have been used as the basis for AppPAL.
We chose to use SecPAL as the basis, however, as it was more lightweight, designed to be readable, and simpler to extend with additional constraints than XACML.

@@

In \autoref{lst:corporate_1} Alice lets Emma specify whether an \code{App} (a variable) \code{isRunnable};
she allows her to delegate the decision (\code{can-say}\/~\code{inf}).
Emma specifies her concerns as policies to be met in \autoref{lst:corporate_2}:
  if Emma is convinced that these are met then she will say the \code{App isRunnable}.
In \autoref{lst:corporate_3} and \autoref{lst:corporate_4} Emma specifies that an app meets the \code{reputable-policy} if the \code{App isBuyable};
  with \code{"google-play"} deciding of what is buyable or not.
Google is not allowed to delegate the decision further, i.e.~Google is not allowed to specify Amazon as a supplier of apps as well.
Emma specifies the \code{"anti-virus-policy"} in \autoref{lst:corporate_5} using a constraint.
When checking the policy the \code{mcAfeeVirusCheck} should be run on the \code{App}.
Only if this returns \code{false} will the policy be met.
To specify the \code{"no-tracking-policy"} Emma says that the \code{'no-location-permissions'} rules implement the \code{'no-tracking-policy'} (\autoref{lst:corporate_6}).
Emma specifies this in \autoref{lst:corporate_7} by checking the app is missing two permissions.

Alice wants to install a new app (\code{com.facebook.katana}) on her phone.
She collects statements to show the app meets the \code{isRunnable} predicate.
\begin{itemize}
  \item\code{"google-play" says "com.facebook.katana" isReputable.}
    Required to convince Emma that the app came from a reputable source.
  \item\code{"emma" says "anti-virus-policy" isMetBy("com.facebook.katana").}
    She can obtain this by running the \ac{av} program on her app.
  \item\code{"emma" says "no-locations-permissions" isMetBy("com.facebook.katana").}
    Needed to show the App meets Emma's no-tracking-policy.
    Emma will say this if the app has no location permissions.
\end{itemize}
These last two statements require the checker to do some extra checks to satisfy the constraints.
To get the second statement AppPAL must run the \ac{av} program on her app and check the result.
The results from the \ac{av} program may change with time as its signatures are updated;
  so the checker must re-run this check every time it wants to obtain the statement connected to the constraint.
For the third statement the AppPAL checker needs to examine the permissions of the app.
It could do this by looking in the \texttt{MANIFEST.xml} inside the app itself, or through the Android package manager if it is running on a device.

@@

Kirin~\cite{Enck:2009ko} is a policy language and tool for enforcing app installation policies to prevent malware.
Policy authors can specify combinations of permissions and broadcast events that should not appear together.
For example, to stop malware sending premium rate text messages, we prevent an app having both the \texttt{SEND\_SMS} and \texttt{WRITE\_SMS} permissions:

\lstinline!restrict permission [SEND_SMS] and permission [WRITE_SMS]!

@@

Future work includes developing and testing, policies for users.
Here we described a policy being specified by a user's employer.
For most end-users writing a policy in a formal language unrealistic.
With Ad-blocking software users subscribe to filter policies written by experts, such as EasyList\footnote{\url{https://easylist.adblockplus.org/en/}}.
We can imagine a similar scheme working well for app installation policies.
Users subscribe to different policies by experts (examples could include no tracking apps, nothing with adult content, no in-app-purchase apps).
Optionally the users could customize the policies further, similar to proof-carrying code~\cite{Necula:1996tr}.

Policy composition raises further questions: what should happen when user's personal and work policies overlap?  
What should happen when policies contradict? 
Future work will look at detecting these problems as well as integrating strategies to resolve them.

Another question might be whether we can use evidence to speed re-checking apps against a policy.
Some static analysis tools, such as Evicheck~\cite{Seghir:2015er}, can create evidence that lets you check an app doesn't have certain behavior faster than it would be to infer the same property in the app without it.
We can also imagine apps being distributed with evidence that proves the app meets an AppPAL policy but avoids the need to check the against the policy explicitly.

@@

AppPAL gives us a framework for describing and evaluating policies for Android apps.
The work provides new, rigorous, ways for machines to enforce user's and device-owner's rules about how apps should behave.
These policies can be enforced more reliably, and with less interaction from person operating the device.

>>>>

There has been much work developing app analysis tools for Android.
Tools such as Stowaway~\cite{Felt:2011kj} detect over-privileged apps.
TaintDroid~\cite{Enck:2010uw} and FlowDroid~\cite{Arzt:2014kf,Li:2015wo} can do taint and control flow analysis; sometimes even between app components.
Other tools like QUIRE~\cite{Bugiel:2012ui} can find privilege escalation attacks between apps.
ScanDAL~\cite{Kim:2012vt} and SCanDroid~\cite{Fuchs:2009vi} help detect privacy leaks.
Appscopy~\cite{Feng:kPGZr_ja} searches for specific kinds of malware.
Tools like DroidRanger~\cite{Zhou:2012tb} scan app markets for malicious apps.
Various tools such as AppGuard~\cite{Backes:2012vm}, Dr.~Android~\&~Mr.~Hide~\cite{Jeon:2012ki} or AppFence~\cite{Hornyack:2011wq} can control the permissions or data an app can get.
MalloDroid~\citep{Fahl:2012dj} looks for apps configured to use SSL incorrectly (for instance by not verifying hostnames or certificates).

@@

Other access control languages, such as XACML 3.0~\cite{Oasis:1M71_oFS}, could also have been used as the basis for AppPAL.
SecPAL however can capture distribution and delegation relationships between
principles and serves as a simplified model of a more complex system like XACML,
and has a well defined semantics, and decidability.

@@

In \autoref{lst:corporate_1} Alice lets Emma specify whether an \code{App} (a variable) \code{isRunnable};
she allows her to delegate the decision (\code{can-say}\/~\code{inf}).
Emma specifies her concerns as policies to be met in \autoref{lst:corporate_2}:
  if Emma is convinced that these are met then she will say the \code{App isRunnable}.
In \autoref{lst:corporate_3} and \autoref{lst:corporate_4} Emma specifies that an app meets the \code{reputable-policy} if the \code{App isBuyable};
  with \code{'google-play'} deciding of what is buyable or not.
Google is not allowed to delegate the decision further, i.e.~Google is not allowed to specify Amazon as a supplier of apps as well.
Emma specifies the \code{'anti-virus-policy'} in \autoref{lst:corporate_5} using a constraint.
When checking the policy the \code{mcAfeeVirusCheck} should be run on the \code{App}.
Only if this returns \code{false} will the policy be met.
To specify the \code{'no-tracking-policy'} Emma says that the \code{'no-location-permissions'} rules implement the \code{'no-tracking-policy'} (\autoref{lst:corporate_6}).
Emma specifies this in \autoref{lst:corporate_7} by checking the app is missing two permissions.

Alice wants to install a new app (\code{com.facebook.katana}) on her phone.
She collects statements to show the app meets the \code{isRunnable} predicate.
\begin{itemize}
  \item\code{'google-play' says 'com.facebook.katana' isReputable.}
    Required to convince Emma that the app came from a reputable source.
  \item\code{'emma' says 'anti-virus-policy' isMetBy('com.facebook.katana').}
    She can obtain this by running the \ac{av} program on her app.
  \item\code{'emma' says 'no-locations-permissions' isMetBy('com.facebook.katana').}
    Needed to show the App meets Emma's no-tracking-policy.
    Emma will say this if the app has no location permissions.
\end{itemize}
These last two statements require the checker to do some extra checks to satisfy the constraints.
To get the second statement AppPAL must run the \ac{av} program on her app and check the result.
The results from the \ac{av} program may change with time as its signatures are updated;
  so the checker must re-run this check every time it wants to obtain the statement connected to the constraint.
For the third statement the AppPAL checker needs to examine the permissions of the app.
It could do this by looking in the \texttt{MANIFEST.xml} inside the app itself, or through the Android package manager if it is running on a device.

@@

Kirin~\cite{Enck:2009ko} is a policy language and tool for enforcing app installation policies to prevent malware.
Policy authors can specify combinations of permissions and broadcast events that should not appear together.
For example, to stop malware sending premium rate text messages, we prevent an
app having both the \texttt{SEND\_SMS} and \texttt{WRITE\_SMS} permissions one
could write: \texttt{restrict permission [SEND\_SMS] and permission [WRITE\_SMS]}.

@@

Future work includes developing and testing, policies for users.
Here we described a policy being specified by a user's employer.
For most end-users writing a policy in a formal language unrealistic.
With Ad-blocking software users subscribe to filter policies written by experts, such as EasyList~\cite{Easylist}.
We can imagine a similar scheme working well for app installation policies.
Users subscribe to different policies by experts (examples could include no tracking apps, nothing with adult content, no in-app-purchase apps).
Optionally the users could customize the policies further.

Policy composition raises further questions: what should happen when user's
personal and work policies overlap or contradict?  
Future work will look at detecting these problems as well as integrating strategies to resolve them.

Another question might be whether we can use evidence to speed re-checking apps against a policy.
Some static analysis tools, such as Evicheck~\cite{Seghir:2015er}, can create evidence that lets you check an app doesn't have certain behavior faster than it would be to infer the same property in the app without it, similar to proof-carrying code~\cite{Necula:1996tr}.
We can also imagine apps being distributed with evidence that proves the app meets an AppPAL policy but avoids the need to check the against the policy explicitly.

@@

AppPAL gives us a framework for describing and evaluating policies for Android apps.
The work provides new, rigorous, ways for machines to enforce user's and device-owner's rules about how apps should behave.
These policies can be enforced more reliably, and with less interaction from the person operating the device.
